---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.path='figure/', warning=FALSE, message=FALSE)
```

## Loading and preprocessing the data

Unzip the data file into a temporary directory, then read data frame, converting date column into Date (NA means default conversion in colClasses).

```{r load}
tmp <- tempdir()
unzip('./activity.zip', exdir = tmp)
df <- read.csv(file.path(tmp, 'activity.csv'), colClasses = c(NA, 'Date', NA))
```

## What is mean total number of steps taken per day?

Calculate and make histogram of total steps per day, excluding NA values.

```{r steps_per_day}
library(ggplot2)
dfDef <- subset(df, !is.na(steps))
stepsPerDay <- aggregate(x = dfDef$steps, by = list(date = dfDef$date), FUN = sum)
qplot(stepsPerDay$x, bins = 20, xlab = 'Steps per day', main = 'Histogram of steps per day')
```

Let's calculate mean and median of total steps per day.

```{r mean_per_day}
# 
meanStepsPerDay <- mean(stepsPerDay$x)
medianStepsPerDay <- median(stepsPerDay$x)
```

The mean of total steps per day is `r meanStepsPerDay` and the median of total steps per day is `r medianStepsPerDay`.


## What is the average daily activity pattern?

Let's create a time series plot of average daily activity.

```{r avg_daily_activity}
avgDailyActivity <- aggregate(x = dfDef$steps, by = list(interval = dfDef$interval), FUN = mean)
qplot(interval, x, 
      data = avgDailyActivity, 
      xlab = 'Daytime', ylab = 'Avg. number of steps', main = 'Daily Activity Graph', geom = 'line')
```

The interval with maximum number of steps

```{r max_daily_number_of_steps}
avgDailyActivity[avgDailyActivity$x == max(avgDailyActivity$x),]$interval
```

## Imputing missing values

The number of rows with NA in steps column. 

```{r rows_with_na}
nrow(df[is.na(df$steps),])
```

Note that no rows with NA in date or interval columns.

```{r na_date}
nrow(df[is.na(df$date),])
nrow(df[is.na(df$interval),])
```
But for example, data for 2012-10-01 contains only NA. 

So, let's replace NA values with averages for the same daily interval.

```{r fill_na}
avgSteps <- avgDailyActivity$x
names(avgSteps) <- avgDailyActivity$interval

dfFilled <- data.frame(df)
for (i in 1:nrow(df)) {
    if (is.na(df[i, 'steps'])) {
        dfFilled[i, 'steps'] <- avgSteps[[as.character(df[i, 'interval'])]]
    }
}
```

Let's make the same histogram as above but for the data frame with imputed values.

```{r hist_imputed_value }

stepsPerDay <- aggregate(x = dfFilled$steps, by = list(date = dfFilled$date), FUN = sum)
qplot(stepsPerDay$x, bins = 20, xlab = 'Steps per day', main = 'Histogram of steps per day w/ imputed values')
meanStepsPerDayImp <- mean(stepsPerDay$x)
medianStepsPerDayImp <- median(stepsPerDay$x)
```

Let's compare mean and median

```{r compare}
meanStepsPerDayImp == meanStepsPerDay
medianStepsPerDayImp == medianStepsPerDay
```

Mean for imputed values `r meanStepsPerDayImp`, mean for NA ignored  `r meanStepsPerDay`.

Median for imputed values `r medianStepsPerDayImp`, median for NA ignored `r medianStepsPerDay`.


## Are there differences in activity patterns between weekdays and weekends?

We can easily see the difference in daily activity from the plot below.

```{r weekdays_daily_activity}
dfFilled$wd <- weekdays(dfFilled$date, abbreviate = TRUE)
dfFilled$day <- rep('weekday')
dfFilled[dfFilled$wd %in% c('Sat', 'Sun'), ]$day <- rep('weekend')
dfFilled$day <- as.factor(dfFilled$day)
avgDailyActivityPerDayType <- aggregate(x = dfFilled$steps, by = list(day = dfFilled$day, interval = dfFilled$interval),  mean)
qplot(interval, x, data = avgDailyActivityPerDayType, facets = day ~ ., geom = 'line', xlab = 'interval', ylab = 'Number of steps', main = 'Avg daily activity per weekdays')
```

